Estimating the number of clusters using Cross Validation

### 2015-06-06 cvclust.tex
Move figures to 'figure' directory
Rename 'myreferences.bib' to 'references.bib'
Add counterexample

@@

After being proposed more than $50$ years, $K$-mean remains one of the most popular and widely used clustering algorithms \citep{jain2010data}. Like many other clustering methods, $K$-mean requires an input parameter $k$, the number of clusters, to be specified by the user. Automatically and quantitatively deciding such parameter is important and yet unsolved problem \citep{fujita2014non}. Various methods have been proposed to tackle this difficulty. One ad hoc approach is to explore the relationship between $W_k$ (within-cluster dispersion) and the number of cluster $k$ for a certain clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases, one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the appropriate number of clusters. The example on the top row of Figure \ref{fig1} demonstrates such approach for data with $k=4$, where the ``elbow" point indeed reveals the true number of clusters. This is based on the idea that under partitioning data set has more impact than over partitioning data set in terms of $W_k$. However, locating the ``elbow" point is somewhat subjective and sometimes is not appropriate to select the optimal $k$. The second example on the bottom row of Figure \ref{fig1} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{Data_example.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{W_k.pdf}
\end{minipage}
\end{minipage}
%\center{\caption{\label{fig1} Left plot shows the actual data, and the right plot shows the $W_k$ \textsl{VS} $k$ curve. }}
%\end{figure}

%\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{Data_elbow.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{elbow.pdf}
\end{minipage}
\end{minipage}
\center{\caption{\label{fig1} Left plots show the actual data, right plots show the corresponding $W_k$ \textsl{VS} $k$ curve. }}
\end{figure}
\\

@@

\bibliography{myreferences}
\bibliographystyle{apalike}
\end{document}

>>>>

\newcommand{\E}{\operatorname{E}}

@@

After being proposed more than $50$ years, $K$-mean remains one of the most popular and widely used clustering algorithms \citep{jain2010data}. Like many other clustering methods, $K$-mean requires an input parameter $k$, the number of clusters, to be specified by the user. Automatically and quantitatively deciding such parameter is important and yet unsolved problem \citep{fujita2014non}. Various methods have been proposed to tackle this difficulty. One ad hoc approach is to explore the relationship between $W_k$ (within-cluster dispersion) and the number of cluster $k$ for a certain clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases, one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the appropriate number of clusters. The example on the top row of Figure \ref{fig1} demonstrates such approach for data with $k=4$, where the ``elbow" point indeed reveals the true number of clusters. This is based on the idea that under partitioning data set has more impact than over partitioning data set in terms of $W_k$. However, locating the ``elbow" point is somewhat subjective and sometimes is not appropriate to select the optimal $k$. The second example on the bottom row of Figure \ref{fig1} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/Data_example.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/W_k.pdf}
\end{minipage}
\end{minipage}
%\center{\caption{\label{fig1} Left plot shows the actual data, and the right plot shows the $W_k$ \textsl{VS} $k$ curve. }}
%\end{figure}

%\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/Data_elbow.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/elbow.pdf}
\end{minipage}
\end{minipage}
\center{\caption{\label{fig1} Left plots show the actual data, right plots show the corresponding $W_k$ \textsl{VS} $k$ curve. }}
\end{figure}
\\

@@

\section{Counterexample with correlated variables}

Suppose that $(X, Y)$ are jointly mean-zero multivariate normal with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  We will show that
Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that
$k = 2$ will be preferred to $k = 1$ almost surely for large enough $n$.

Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  In
$k = 2$ case, then if $n$ is large enough, the cluster assignment will be
determined according to whether $Y > 0$.

In the $k = 1$ case, if $n$ is large enough, the average Gabriel
cross-validation error converges to $\E(Y^2) = 1$. In the $k = 2$ case, if $n$
is large enough, then the linear decision rule for the $X$ variables will be
determined according to whether $X > 0$.  In this case, the error converges to
$\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$, where $a =
\sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows that $Y
\mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$.   Thus, in
the $k = 2$ case, the Gabriel cross-validation error converges to $1 + a^2 (1
- 2 \rho)$.  In particular, if $\rho > 0.5$, then the asymptotic
cross-validation error for $k = 2$ will be smaller than for $k = 1$.

We confirm this with a simulation.  [TODO: add simulation]

\bibliography{references}
\bibliographystyle{apalike}
\end{document}
